---
title: "Monitoring S3 access"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Monitoring S3 Access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  dpi        = 300,
  fig.width  = 12,
  fig.height = 6,
  message    = FALSE,
  warning    = FALSE,
  eval       = FALSE
)
```

Every request to a CORI data bucket is written to an S3 server access log.
This vignette turns those logs into charts: who is downloading what, from
which bucket, and how that changes over time.

Chunks are set `eval = FALSE` because rendering them queries several months
of live log data. Run them interactively against a real window.

## Where the logs live

Server access logging is enabled on every allowlisted bucket except
`cori.data.verse`, which is the destination. Log objects land under a
date-partitioned prefix that encodes both the source bucket and the day:

```
s3://cori.data.verse/logs/312512371189/us-east-1/<bucket>/YYYY/MM/DD/<file>
```

That layout matters for performance. Narrowing the glob to the months you
care about is what keeps a query reading megabytes instead of the full
history — there are hundreds of thousands of small log objects.

```{r libraries}
library(cori.data.s3)
library(cori.charts)
library(dplyr)
library(ggplot2)
library(scales)

load_fonts()
```

## Reading the logs

Access logs are space-delimited with a bracketed timestamp and a quoted
request URI, so they are parsed with a single regular expression rather
than a CSV reader. The important discipline is to aggregate inside DuckDB
and return only a small summary to R — never pull raw log lines into
memory.

```{r read}
con <- connect_to_s3("cori.data.verse")
on.exit(DBI::dbDisconnect(con, shutdown = TRUE), add = TRUE)

# Widen or narrow this glob to set the reporting window.
glob <- "s3://cori.data.verse/logs/312512371189/us-east-1/*/2026/*/*/*"

log_pattern <- paste0(
  "^(\\S+) (\\S+) \\[([^\\]]+)\\] (\\S+) (\\S+) (\\S+) (\\S+) (\\S+) ",
  "\"[^\"]*\" (\\S+) (\\S+) (\\S+)"
)

activity <- DBI::dbGetQuery(con, sprintf("
  WITH raw AS (
    SELECT regexp_extract(line, '%s',
      ['owner','bucket','ts','ip','requester','reqid',
       'operation','key','status','err','bytes']) AS f
    FROM read_csv('%s', delim='\\x01', header = false,
                  columns = {'line': 'VARCHAR'})
  )
  SELECT
    f.bucket                                                   AS bucket,
    CAST(strptime(split_part(f.ts, ' ', 1),
                  '%%d/%%b/%%Y:%%H:%%M:%%S') AS DATE)          AS day,
    CASE
      WHEN f.requester = '-' THEN 'anonymous'
      WHEN f.requester LIKE '%%CoriDataS3ReaderRole%%'
        THEN coalesce(nullif(regexp_extract(
               split_part(f.requester, '/', 3),
               '^coridata-(.*)-[0-9]+$', 1), ''), 'anon')
      ELSE 'local credentials'
    END                                                        AS caller,
    COUNT(*)                                                   AS requests,
    SUM(TRY_CAST(f.bytes AS BIGINT))                           AS bytes
  FROM raw
  WHERE f.operation LIKE 'REST.GET.OBJECT%%'
  GROUP BY 1, 2, 3
", log_pattern, glob))

glimpse(activity)
```

One row per bucket, day, and caller. Everything below is `dplyr` on that
frame — the expensive work is already done.

## Downloads over time, by bucket

```{r chart-by-bucket}
activity |>
  group_by(bucket, day) |>
  summarise(requests = sum(requests), .groups = "drop") |>
  ggplot(aes(day, requests, color = bucket)) +
  geom_line(linewidth = 0.8) +
  scale_y_continuous(labels = label_comma()) +
  scale_x_date(date_labels = "%b %Y") +
  scale_color_cori() +
  labs(
    title    = "S3 object downloads by bucket",
    subtitle = "Daily GET requests recorded in server access logs",
    x = NULL, y = "Requests", color = NULL
  ) +
  theme_cori_line()
```

## Who is downloading

The `requester` field carries the full assumed-role session ARN, and the
session name is minted per credential as `coridata-<caller>-<timestamp>`.
That is what the query above unpacks into `caller`. Requests with no
credentials appear as `anonymous`, which on the seven public buckets is
usually the bulk of the traffic.

```{r chart-by-caller}
activity |>
  group_by(caller) |>
  summarise(requests = sum(requests), .groups = "drop") |>
  slice_max(requests, n = 12) |>
  ggplot(aes(reorder(caller, requests), requests)) +
  geom_col(fill = cori_colors[["Mid Teal"]]) +
  coord_flip() +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title    = "Downloads by caller",
    subtitle = "Parsed from the vended credential's session name",
    x = NULL, y = "Requests"
  ) +
  theme_cori_horizontal_bars()
```

## Year-to-date summary

```{r ytd}
activity |>
  filter(day >= as.Date(format(Sys.Date(), "%Y-01-01"))) |>
  group_by(bucket) |>
  summarise(
    requests = sum(requests),
    gb       = round(sum(bytes, na.rm = TRUE) / 1024^3, 1),
    callers  = n_distinct(caller[!caller %in% c("anonymous", "local credentials")]),
    .groups  = "drop"
  ) |>
  arrange(desc(requests)) |>
  knitr::kable()
```

Swap the `filter()` for `day >= Sys.Date() - 90` to get the trailing
ninety days instead.

## Caveats

Server access logging is best effort. Most requests produce a record and
most records arrive within a few hours, but delivery is neither guaranteed
nor provably complete, and latency means today's activity is not visible
yet. These numbers are a reliable picture of usage, not an audit trail —
do not present them as a compliance artifact.

Callers appear as `anon` unless the calling code passes a `caller` tag:

```{r caller-tag}
con <- connect_to_s3("cori.data.qcew", caller = "quarterly-refresh")
```
