---
title: "Monitoring S3 access"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Monitoring S3 access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  dpi        = 300,
  fig.width  = 12,
  fig.height = 6,
  message    = FALSE,
  warning    = FALSE
)
```

Every request to a CORI data bucket is written to an S3 server access log.
This vignette turns those logs into charts: who is downloading what, from
which bucket, and how that changes over time.

Chunks are set `eval = FALSE` because rendering them queries several months
of live log data. Run them interactively against a real window.

## Where the logs live

Server access logging is enabled on every allowlisted bucket except
`cori.data.verse`, which is the destination. Log objects land under a
date-partitioned prefix that encodes both the source bucket and the day:

```
s3://cori.data.verse/logs/312512371189/us-east-1/<bucket>/YYYY/MM/DD/<file>
```

That layout matters for performance. Narrowing the glob to the months you
care about is what keeps a query reading megabytes instead of the full
history — there are hundreds of thousands of small log objects.

```{r libraries, eval = FALSE}
library(cori.data.s3)
library(cori.charts)
library(dplyr)
library(purrr)
library(ggplot2)
library(scales)

load_fonts()
```

## Reading the logs

Access logs are space-delimited with a bracketed timestamp and a quoted
request URI, so they are parsed with a single regular expression rather
than a CSV reader. The important discipline is to aggregate inside DuckDB
and return only a small summary to R — never pull raw log lines into
memory.

```{r read, eval = FALSE}
# One glob per day for the past week — each query hits a single date prefix.
dates <- seq(Sys.Date() - 7, Sys.Date(), by = "day")
globs <- sprintf(
  "s3://cori.data.verse/logs/312512371189/us-east-1/*/%s/*",
  format(dates, "%Y/%m/%d")
)

log_pattern <- paste0(
  "^(\\S+) (\\S+) \\[([^\\]]+)\\] (\\S+) (\\S+) (\\S+) (\\S+) (\\S+) ",
  "\"[^\"]*\" (\\S+) (\\S+) (\\S+)"
)

query_template <- "
  WITH raw AS (
    SELECT regexp_extract(line, '%s',
      ['owner','bucket','ts','ip','requester','reqid',
       'operation','key','status','err','bytes']) AS f
    FROM read_csv('%s', delim='\\x01', header = false,
                  columns = {'line': 'VARCHAR'})
  )
  SELECT
    f.bucket                                                   AS bucket,
    CAST(strptime(split_part(f.ts, ' ', 1),
                  '%%d/%%b/%%Y:%%H:%%M:%%S') AS DATE)          AS day,
    CASE
      WHEN f.requester = '-' THEN 'anonymous'
      WHEN f.requester LIKE '%%CoriDataS3ReaderRole%%'
        THEN coalesce(nullif(nullif(regexp_extract(
               split_part(f.requester, '/', 3),
               '^coridata-(.*)-[0-9]+$', 1), ''), 'anon'), 'anonymous')
      ELSE 'local credentials'
    END                                                        AS caller,
    COUNT(*)                                                   AS requests,
    SUM(TRY_CAST(f.bytes AS BIGINT))                           AS bytes
  FROM raw
  WHERE f.operation LIKE 'REST.GET.OBJECT%%'
  GROUP BY 1, 2, 3
"

con <- connect_to_s3("cori.data.verse")

activity <- purrr::map(globs, \(g) {
  DBI::dbGetQuery(con, sprintf(query_template, log_pattern, g))
}) |>
  dplyr::bind_rows() |>
  dplyr::filter(
    bucket %in% c(
      "cori.data.bds",
      "cori.data.bfs",
      "cori.data.bps",
      "cori.data.fcc",
      "cori.data.hu",
      "cori.data.pep",
      "cori.data.qcew",
      "ruraldefinitions"
    )
  )

glimpse(activity)
```

One row per bucket, day, and caller. Everything below is `dplyr` on that
frame — the expensive work is already done.

## Downloads over time, by bucket

```{r chart-by-bucket, eval = FALSE}
activity |>
  filter(day >= Sys.Date() - 7) |>
  group_by(bucket, day) |>
  summarise(requests = sum(requests), .groups = "drop") |>
  ggplot(aes(day, requests, color = bucket)) +
  geom_line(linewidth = 0.8) +
  scale_y_continuous(labels = label_comma()) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 day") +
  scale_color_viridis_d(option = "turbo", begin = 0.1, end = 0.9) +
  labs(
    title    = "S3 object downloads by bucket",
    subtitle = "Daily GET requests over the past week",
    x = NULL, y = "Requests", color = NULL
  ) +
  theme_cori_line()
```

## Who is downloading

The `requester` field carries the full assumed-role session ARN, and the
session name is minted per credential as `coridata-<caller>-<timestamp>`.
That is what the query above unpacks into `caller`.

`anonymous` deliberately combines two things that are technically different
but mean the same thing for this report: a request with no credentials at
all, and a vended credential nobody tagged with a `caller`. Both answer "who
did this?" with nothing. That's the question this report exists to answer,
so the two collapse into one number rather than being split by a technical
distinction (had a credential vs. didn't) that doesn't matter here. On the
seven public buckets, `anonymous` is usually the bulk of the traffic.
`local credentials` is a separate, real category: a known AWS identity that
simply didn't go through the vending endpoint.

```{r chart-by-caller, eval = FALSE}
activity |>
  group_by(caller) |>
  summarise(requests = sum(requests), .groups = "drop") |>
  slice_max(requests, n = 12) |>
  ggplot(aes(reorder(caller, requests), requests)) +
  geom_col(fill = cori_colors[["Mid Teal"]]) +
  # Centered at each bar's own midpoint (requests / 2), not inset from the
  # tip -- real traffic here is heavily skewed (anonymous vastly outweighs
  # everything else), and a fixed tip-inset would overflow a short bar.
  # Centering scales with each bar's own length regardless of skew.
  geom_text(
    aes(y = requests / 2, label = scales::comma(requests)),
    color    = "white",
    fontface = "bold",
    size     = 3.5
  ) +
  coord_flip() +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title    = "Downloads by credential type",
    subtitle = "Parsed from the vended credential's session name",
    x = NULL, y = "Requests"
  ) +
  theme_cori_horizontal_bars()
```

## Year-to-date summary

```{r ytd, eval = FALSE}
activity |>
  filter(day >= as.Date(format(Sys.Date(), "%Y-01-01"))) |>
  group_by(bucket) |>
  summarise(
    requests = sum(requests),
    gb       = round(sum(bytes, na.rm = TRUE) / 1024^3, 1),
    callers  = n_distinct(caller),
    .groups  = "drop"
  ) |>
  arrange(desc(requests)) |>
  knitr::kable()
```

Swap the `filter()` for `day >= Sys.Date() - 90` to get the trailing
ninety days instead.

## Caveats

Server access logging is best effort. Most requests produce a record and
most records arrive within a few hours, but delivery is neither guaranteed
nor provably complete, and latency means today's activity is not visible
yet. These numbers are a reliable picture of usage, not an audit trail —
do not present them as a compliance artifact.

In the raw log data, a vended session with no `caller` tag literally
embeds `anon` in its session name (that's the vending endpoint's own
fallback) -- which is why the query above folds it into `anonymous` rather
than treating "vended but untagged" as its own category. Passing a tag is
what moves a request out of `anonymous` and into an attributable name:

```{r caller-tag, eval = FALSE}
con <- connect_to_s3("cori.data.qcew", caller = "quarterly-refresh")
DBI::dbDisconnect(con, shutdown = TRUE)
```
